---
title: "chapter5.Rmd"
author: "Rstudio"
date: "24 helmikuuta 2017"
output: html_document
---

#RStudio Exercise 5

```{r, message=FALSE, warning=FALSE, echo=FALSE}
setwd("Z:/Jatko-opinnot/Tilastotiede 2016/IODS/IODS-project/Data")
human <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep=",", header=T)
```

```{r, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
#install and use the package 
install.packages("GGally", repos="http://cran.rstudio.com/")
library(GGally)
```

```{r, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
#install and access the package 
install.packages("ggplot2", repos="http://cran.rstudio.com/")
library(ggplot2)
```

##Describing the dataset

The 'human' dataset originates from the United Nations Development Program. More information about the dataset can be found here: http://hdr.undp.org/en/content/human-development-index-hdi. In this assignment, I use the dataset which has 8 variables and 155 countries. The variables of the dataset describe health, economic and gender equality of each country. Below is the list of the variable names and descriptions. 

> 
* "GNI" = Gross National Income per capita
* "Life.Exp" = Life expectancy at birth
* "Edu.Exp" = Expected years of schooling 
* "Mat.Mor" = Maternal mortality ratio
* "Ado.Birth" = Adolescent birth rate
* "Parli.F" = Percetange of female representatives in parliament
* "Edu2.FM" = the ratio of Female and Male populations with secondary education in each country
* "Labo.FM" = the ratio of labour force participation of females and males in each country

Below is the summary of the data. We see that the variables have very different scales. The differences between the scales are important for later analyses.

```{r}
summary(human)
```

Just for curiosity, here is how Finland is doing

```{r}
human[22,]
```

And some other countries as well. As expected, there are huge differences between countries.
```{r}
head(human, n=5)
tail(human, n=5)
```

```{r}
p <- ggpairs(human, mapping = aes(alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

In the picture above we have the graphical overview of the dataset, showing the distributions and correlations of the variables. The variables do not seem to be normally distributed (the variable Edu.Exp, which describes the expected years of schooling is closes to normal distribution). The variables describing health are strongly correlated with each other: for instance, maternal mortality ratio ("Mat.Mor"), has strong, inverse correlation with life expectancy at birth (r=-0.86) and adolescent birth rate (r:-0.76). Maternal mortality is also associated with the access to education: it has an inverse association with the expected years of schooling (r=-0.74) and the ratio of female and male populations with secondary education (r=-0.66). Gross National Income per capita, GNI, is positively associated with life expectancy at birth (r=0.63) and expected years of schooling (r=0.62). GNI is also at least moderately associated with the ratio of female and male populations with secondary education (r=0.43), but - interestingly - GNI is not correlated with all of the variables describing gender equality. GNI is not associated with the ratio of labor force participation of females and males as well as with percentage of female representatives in parliament.

##Principal Component Analysis 
###Non-standardized data
Next I perform principal component analysis (PCA) on the not standardized human data. My goal is to find the components that capture the most of the variance. PCA can be performed by two slightly different matrix decomposition methods from linear algebra: the Eigenvalue Decomposition and the Singular Value Decomposition (SVD). In this assignment, I used the SVD method. I use the prcomp() function which decompose a data matrix into a product of smaller matrices, which let's us extract the underlying principal components. This makes it possible to approximate a lower dimensional representation of the data by choosing only a few principal components.

```{r}
pca_human <- prcomp(human)
pca_human
```

In PCA, the data is first transformed to a new space with equal or less number of dimensions. These new features are called the principal components. The 1st principal component (PC1) captures the maximum amount of variance from the features in the original data. The 2nd principal component (PC2) is orthogonal to the first and it captures the maximum amount of variability left. The same is true for each principal component. They are all uncorrelated and each is less important than the previous one, in terms of captured variance. With our data, we found 8 principal components. Lets see the importance of different components.

```{r}
summary(pca_human)
```


```{r, message=FALSE, warning=FALSE, results='hide'}
# create and print out a summary of pca_human
s <- summary(pca_human)
# rounded percetanges of variance captured by each PC
pca_pr <- round(100*s$importance[2, ], digits = 1)
# create object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)") 
```

```{r, message=FALSE, warning=FALSE}
# draw a biplot
biplot(pca_human, cex = c(0.5, 0.8), xlab = pc_lab[1], ylab = pc_lab[2])
```

This biplot suggests that almost all of the variance is caused by one GNI, Gross National Income per capita. PCA is sensitive to the relative scaling of the original features and assumes that features with larger variance are more important than features with smaller variance. GNI has the largest scale (min: 581, max: 123,100). Therefore I'll do the analysis again, but with this time by standardized data.

###Standardized data

```{r}
#scale the dataset
human_st <- scale(human)
#Pca on standardized data
pca_human_st <- prcomp(human_st)
pca_human_st
summary(pca_human_st)
```

The results of the PCA with standardized data are clearly different. Now the first two components explain 53.6% and 16.2% of the variance in our dataset. The variables having strongest loadings on PC1 are maternal mortality ratio and adolescent birth rate. The strongest negative loadings on PC1 are life expectancy, expected years of schooling, the ratio of females to males with secondary education, and gross national income per capita. The variables having strongest loadings on PC2 are the ratio of females to males in labor and female parliamentary representation. The first principal component seems to describe health and economic factors of the countries, and the second describes the gender equality in participating in politics and labor. Together these two component explained 70 percentages of the variance between countries.

```{r, message=FALSE, warning=FALSE, results='hide'}
s_td <- summary(pca_human_st)
# rounded percetanges of variance captured by each PC
pca_pr2 <- round(100*s_td$importance[2, ], digits = 1)
# create object pc_lab to be used as axis labels
pc_lab2 <- paste0(names(pca_pr2), "(", pca_pr2, "%)")   
```


```{r, message=FALSE, warning=FALSE}
# draw a biplot
biplot(pca_human_st, cex = c(0.5, 0.8), col = c("grey40", "red"), xlab = pc_lab2[1], ylab = pc_lab2[2])
```

From the biplot we see that, for instance, Iceland has high gender equality, giving birth is safe to a woman's health, and women tend to participate in politics and labor force more than most of the countries. Many African and Asian countries such as Niger and Afghanistan, for instance, are poor countries where people have short life expectancy, women tend to have many children and giving birth is dangerous, and women do not participate in politics and labor. Rwanda seems an interesting case since they have women participating in politics and labor force but it is still a poor country relatively high maternal mortality and adolescent birth rate. Qatar, Kuwait and Saudi Arabia are among the wealthiest countries in the world but women do not participate in the labor force or politics. This biplot clearly states that if we want to measure well-being of populations of different countries, many measures are needed. Also, gender inequality is not automatically improving when gross national income per capita improves.

##Multiple Correspondence Analysis

```{r, message=FALSE, warning=FALSE, echo=FALSE}
install.packages("FactoMineR", repos="http://cran.rstudio.com/")
library(FactoMineR)
library(dplyr)
install.packages("tidyr", repos="http://cran.rstudio.com/")
library(tidyr)
data(tea)
```
I installed the packages FactoMineR and here is the dataset "tea".


```{r}
dim(tea)
str(tea)
```

I select six variables for a new dataset and draw a graph. My new dataset consists of six variables describing how one does drink tea, what kind of tea, does one drink tea during lunch, does one drink tea with sugar, where one buys his or her tea, and the gender of the person.

```{r}
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "sugar", "where", "lunch", "sex")
# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, one_of(keep_columns))
# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + facet_wrap("key", scales = "free")
```

Next I'll do multiple correspondence analysis (MCA) which is a method to analyze qualitative data and can be used to detect patterns or structure in the data as well as in dimension reduction.

```{r}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)
# summary of the model
summary(mca)
#draw a plot
plot(mca, invisible=c("ind"), habillage = "quali")
```

Below is the graph of the results of MCA. The distance between variable categories gives a measure of their similarity. 

```{r, message=FALSE, warning=FALSE, results='hide'}
s <- summary(mca)
# rounded percetanges of variance captured by each PC
s2 <- round(100*s$importance[2, ], digits = 1)
# create object pc_lab to be used as axis labels
s2_lab <- paste0(names(s2), "(", s2, "%)") 